---
title: "workspace"
runtime: shiny
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo= FALSE}
  library(ggplot2)
  library(extracat)
  library(tidyverse)
  library(shiny)
  library(GeomMLBStadiums)
```


After loading in the pbp2019 data set that will be used throughout this analysis, we will first examine the missing values. The data set has too many features to make use of a graph, but a few clear trends exist after looking at the column missing values and taking a look at the data. Many variables such as pitcher, batter, strikes, and balls, have no missing values, while others have large amounts. The explanation for this is inuitive: each pitch represents a row of data and the values without missing values are the values that exist no matter the outcome of the pitch. While many of the variables missing many values depend on the outcome of the pitch. A couple examples: launch_angle will be missing if the ball was not hit. on_2b will be missing if no one is on second base.

Interestingly, we see related variables share the same total count of missing variables, such as all of the fielder_ variables having a missing value of 3159. The best explanation of this is that some ballparks (games in different countries) may not have measured these variables. 

After looking at the colMeans, we see that a few variables are missing for every value in the dataset, so we drop them from the data frame. 

```{r}
load('pbp2019.rda')
pbp2019[pbp2019=='null']<- NA

colSums(is.na(pbp2019)) %>% sort(decreasing=TRUE)

colMeans(is.na(pbp2019))

drop <- c('spin_dir', 'spin_rate_deprecated','break_angle_deprecated','break_length_deprecated','tfs_deprecated','tfs_zulu_deprecated','umpire') # columns missing every value
pbp2019 <- pbp2019[,!(names(pbp2019) %in% drop)] # drop useless columns

```


The season data frame has no missing values, as it is an aggregated data set over the season.


```{r} 
# read in season data frame
season_data <- read.csv('FanGraphs_Leaderboard.csv')

colnames(season_data)[1] <- 'Season'
season_data[season_data=='null']<- NA
colSums(is.na(season_data)) %>% sort(decreasing=TRUE)
```


An often noted trend in the MLB has been the increase in home runs, so we will look at a histogram faceted by season to see if teams tended to hit more home runs in recent years than previously. Clearly, this trend is noted for a reason as the histogram dramatically shifts over the years.  

```{r}
ggplot(season_data)+
  geom_histogram(aes(x = HR), fill = "lightBlue", col = "black") +
  ggtitle("Homeruns each season")+
  facet_wrap(~Season)

```


WAR is a new-age baseball statistic that is widely regarded as being the most comprehensive measure of a player or teams success. For this reason, the following graph shows each teams total WAR for a season over the last 10 seasons. Basically, the graph exhibits how consistent each team has been (the spread of data), as well as how successful the team has been (the further right the histogram is the more successful). A few examples: the Rockies have been consistent, but not very successful, the Astros have been very inconsistent, but had some very successful years in this period. 

```{r}
ggplot(season_data)+
  geom_histogram(aes(x = WAR), binwidth=3, fill = "lightBlue", col = "black") +
  ggtitle("WAR by team over time")+
  facet_wrap(~Team)
```


Pitchers throw different pitches for different reasons. Some pitches move more than others, while others are thrown with more speed. Therefore, we wanted to compare the pitch speed based on different type of pitches. The first observation is that some pitches occur with a much higher frequency than others. As expected, the pitches create normal distributions with the means of each pitch varing greatly by pitch. Four-seam fastballs (FF) are known for being the most utilized and fastest pitch, which the chart verifies. Additionally, some pitches such as sliders (SL) have a wider range of speeds with which they are thrown. For baseball fans, this is also to be expected as sliders are harder to classify and come in many forms. 


```{r echo=FALSE}
pbp2019 <- transform(pbp2019, plate_x = as.numeric(plate_x), plate_z = as.numeric(plate_z), release_speed = as.numeric(release_speed), hc_x = as.numeric(hc_x),hc_y = as.numeric(hc_y), launch_speed = as.numeric(launch_speed)) # correcting column formats

empty_pt <- c('','EP','FO','KN',NA) # the following pitches do not have enough data to make meaningful analysis, so we will remove

temp_pt <- pbp2019[! pbp2019$pitch_type %in% empty_pt,] # remove the above pitches 

ggplot(temp_pt)+
  geom_histogram(aes(x= release_speed), binwidth=1, fill = "lightBlue", col = "black") +
  ggtitle("Pitch speed by type")+
  facet_wrap(~pitch_type)

```

Going back to the home run trend, as it has been the talk of baseball for the past couple of years. Looking at how each teams home run totals have changed over the past few years is useful, but as can be seen in the graph below, including all 32 teams is a little crowded. 

```{r}
ggplot(season_data, aes(Season, HR, color = Team)) + geom_line() +
    ggtitle("Home runs over time") +
    labs (x = "Season", y = "Total home runs") +
    scale_x_continuous( breaks= c(2010,2015,2019))+
    theme_grey(16) +
    theme(legend.title = element_blank())

```

Therefore, we will look at comparing the Yankees and Twins, the two teams who took part in an incredible home run race last year to see how both teams have changed over the past 10 years. Clearly the Yankees have shown more consistent home run ability than the Twins. 

```{r}
hr_race <- c('Yankees', 'Twins')
temp_hr <- season_data[season_data$Team %in% hr_race,]

ggplot(temp_hr, aes(Season, HR, color = Team)) + geom_line() +
    ggtitle("Home runs over time NYY vs. MIN") +
    labs (x = "Season", y = "Total home runs") +
    scale_x_continuous( breaks= c(2010,2015,2019))+
    theme_grey(16) +
    theme(legend.title = element_blank())


```


Comparing WAR for teams from the NL East will give us a great idea of which teams in the division have been consistent and which have been successful. The graph creates an interesting way to track the trajectory of each team over the past decade. The Phillies were very successful early in the decade, only to fall off greatly in the middle parts, and begin their rise back up over the last couple of years. Another interesting note is that the Phillies were second to last in the division in 2019 and went on to win the World Series. 

```{r}
nl_east <- c('Marlins','Mets','Nationals','Phillies','Braves')
temp_nl <- season_data[season_data$Team %in% nl_east,]


ggplot(temp_nl, aes(Season, WAR, color = Team)) + geom_line() +
    ggtitle("WAR over time NL East") +
    labs (x = "Season", y = "Total WAR") +
    scale_x_continuous( breaks= c(2010,2015,2019))+
    theme_grey(16) +
    theme(legend.title = element_blank())

```

Heat maps have an obvious application to baseball in terms of the strike zone. Many different analysis can be done based off pitch location, such as hit type, swing percentage, etc. We are going to take the approach of comparing two batters heat maps based on whether the pitch was a ball, strike, or put into play. Ideally, a batter would like to have a tightly bound cluster of pitch location put into play, with a small outer ring of swinging strikes, and anything not in the strike zone watched for a ball. 

To analyze, we chose one batter known for having a good eye of the strike zone (Mike Trout) and one known for swinging at anything (Lewis Brinson). The graph shows the expected result, that Trout (545361) has a much tighter zone of strikes/balls in play than Brinson (621446). Trout puts more balls in play within the zone and swings at less balls outside the zone. 

```{r}
trout_brinson <- pbp2019 %>%
  filter(batter == 545361 | batter == 621446)


ggplot(trout_brinson)+
  geom_bin2d(aes(plate_x, plate_z, fill = type))+
  facet_grid(~batter)

```


To continue comparing the teams within the NL East, we read in a data set from fangraphs.com that is just the aggregate of the season_stats we have been using for each team. We wanted to compare how each of the teams compare across each statistical category. The analysis provides a comprehensive view about the categories each team has performed well in against the other teams in the division. 

A few clear insights appear. First, the Nationals are the best in almost every category, except for Def (defensive rating), where the Braves have been the best over the years. Secondly, the Marlins have been worst in just about every category except for BABIP (batting average on balls in play) which is known to be the most random statistic in baseball. The performance of the teams over the decade seem to make a clear order based on the chart: Nationals > Braves/Mets > Phillies > Marlins. 

```{r}
library(viridis)
# from before nl_east <- c('Marlins','Mets','Nationals','Phillies','Braves')
agg_season_data <- read.csv('agg_leaderboard.csv')
temp_agg_nl <- agg_season_data[agg_season_data$Team %in% nl_east,]

row.names(temp_agg_nl) <- temp_agg_nl$Team
temp_agg_nl$Team <- NULL

season_scaled <- data.frame(scale(temp_agg_nl)) %>%
  rownames_to_column() %>%
  gather(colname, value, -rowname)


ggplot(season_scaled, aes(x = rowname, y = colname, fill = value)) +
  geom_tile() + scale_fill_viridis() +
  ggtitle("Team Stats scaled")

```


As outlined in Luke and Chris' community contribution project, the package GeomMLBStadiums can be used to create a spray chart for balls in play. We used a Shiny Application to make this tool as flexible as the user would like it to be. The app provides the opportunity to choose which batter to analyze, while restricting the pitch type to certain pitches and certain areas of the strike zone. The horizontal and vertical pitch location allow the user to restrict the pitches to certain areas. So questions of the following sort can be answered: "how does Christian Yelich hit pitches that are high and outside compared to pitches low and inside?" Enjoy the tool!


```{r echo = FALSE}
selectInput("p_type", label = "Pitch Type: FF= Four-seam FB, SL=Slider, SI= Sinker, etc.",
              choices = c(c("FF"),c("FS"),  c("SL"),c("SI"),c("CH"),c("FC"),c("CU"),c("FT"),c("KC")), multiple=TRUE, selected = c("FF"))

sliderInput("x_axis", label = "Horizontal pitch location (0 is middle of the plate, strike zone ~ [-1, 1]):",
              min = -2 , max = 2, value = c(-2,2), step = 0.05)
sliderInput("y_axis", label = "Vertical pitch location (strike zone ~ [2,4] :",
              min = 0 , max = 5, value = c(0,5), step = 0.05)

textInput("batter", label = "Batter ID # (found on MLB.com)- ex. Mike Trout = 545361", value = "545361")
```

```{r echo = FALSE}

renderPlot({
  pbp2019 %>%
  filter(pitch_type %in% input$p_type,batter == as.numeric(as.character(input$batter)), plate_x > input$x_axis[1], plate_x < input$x_axis[2], plate_z > input$y_axis[1], plate_z <input$y_axis[2]) %>% mlbam_xy_transformation() %>%  
  ggplot(aes(x=hc_x_, y=hc_y_, color = bb_type )) + 
  geom_spraychart(stadium_ids = "brewers",
                  stadium_transform_coords = TRUE, 
                  stadium_segments = "all") + 
  theme_void() + 
  coord_fixed()

})
```
